Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning

### 2020-05-25 paper/main.tex
dynamic supplement and stuff

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Manuscript Category}
\usepackage{url}
\usepackage{placeins}

@@

\subsection{Data Pre-processing}
We first pre-process and store the raw genomic information by using GeenuFF 
\citep{denton2019}. GeenuFF is a tool for checking and exploring genomic data, that 
stores all information inside a SQL database. Training and evaluation-ready data is 
generated by querying this database and then transforming the returned data into a 
numerical format suitable for machine learning. The encoding of the genomic sequence 
is done in line with the IUPAC nucleic acid notation and the functional annotation 
used as labels during training is transformed to a one hot encoding with the four 
classes {\it Intergenic}, {\it UTR}, {\it CDS} and {\it Intron}. See also Table 
\ref{tab:arrays} for a more detailed description of the generated data types.

@@

Our final model was trained with 10 bases of genomic sequence as input during each 
time step and produces individual predictions for each of those 10 bases 
simultaneously. This grouping enabled us to train effectively with far longer 
sequences than usual. 
We compare our final models to a dilated CNN (dCNN) baseline as this kind of architecture is 
widely used when working with DNA data as input. Details of the neural architecture 
search are given in Section \ref{sec:dcnn} of the Supplements.

\subsection{Inference Techniques}
We also used multiple techniques to improve the prediction quality after the 
training was done. One very effective way for genomes with larger genes was to input 
longer sequences during inference time than during training. This was done for all 
animal genomes except the invertebrates. The input sequences were up to 10 times 
longer, depending on the phylogenetic group and assembly quality. This also demonstrates the 
ability of our model to generalize as it is able to make successful prediction on 
far longer sequences than it has ever seen. The 
concrete lengths were chosen to keep the typical average gene length roughly 
proportional to the length of the sequence input. For more implementation details 
on this see Section \ref{sec:longer} of the Supplements or the source code.

The final predictions of a single model are constructed by overlapping predictions, 
that were made from a sliding window and then cropped to a core sequence. This was done 
to strongly reduce a typical drop in performance of the models towards the beginning 
and end of each sequence (see Figure \ref{fig:main_results}c). It also improves the 
average model performance in general by providing the model with multiple different 
starting points. The different overlapping sequences were combined by averaging 
the individual softmax values of each base. The figures in Section \ref{sec:overlapping} of the 
Supplements show the effect of overlapping for each genome, ordered by N75. We found 
that overlapping tends to work best if the genomes are not very fragmented and we 
used it for both animals and plants.

@@

\subsection{Evaluation of AUGUSTUS} 
\label{sec:augustus}
Due to the practical time constraints for retraining and running, we decided to compare our Helixer
models only to one existing {\it de novo} gene caller, the popular gene caller, Augustus \citep{stanke2003gene}.
Typical usage requires retraining Augustus for each species, with the exception of a
few lucky cases where a model for a sufficiently close relative is already available. To scale this for 
the large plant and animal datasets, we used protein-homology to create a training set and
therefore could train and evaluate only models without UTRs.

Orthologs of highly conserved generally single-copy genes were identified in each genome using
BUSCO. The {\it viridiplantae} set was used for plants and the {\it metazoa} set for animals.
The training genbank files were generated directly by BUSCO, by utilizing the ``--long"  % inline code?
parameter. For plants, the entire retraining could be performed as above; however, for animals
we randomly selected only half the BUSCO-generated training set, which resulted in a training set 
size and runtime comparable to the plants (~400 genes and several days per species). For animals
the training was carried out with the subsetted training file and using the following scripts provided
by Augustus. An untrained model was setup with "new\_species.pl", and the model was fit by running
"etraining" before and after the major hyper-parameter optimization with "optimize\_augustus.pl".
Using the trained model for each species, we ran the main prediction ("augustus") with ``--UTR=off"
and ``--gff3=on". The gffs produced by Augustus were imported into GeenuFF and exported as h5
files in the same manor as the reference, allowing direct comparison with reference and 
Helixer predictions. While tenable for some 237 species, this method allowed neither training
nor prediction of UTR regions with Augustus, so the metric Subgenic F1, which excludes UTR,
was used for all comparisons between Augustus and Helixer.

\subsection{Evaluation against independent RNAseq data}
\label{sec:rnaseq}
The reference annotations were created with existing tools and largely with a pipeline incorporating
{\it de novo} gene predictions with RNAseq and homology data. The references, like any
data, are expected to contain errors so we chose to use RNAseq data for independent evaluation.
In plants there is the additional concern that the references may share biases with
Augustus or HMMs as Augustus was used as the {\it de novo} gene caller for the reference of
many plant species (e.g. {\it Arabidopsis lyrata}, \cite{rawat2015improving}; 
{\it Asparagus officinalis}, 
\cite{harkess2017asparagus}; and {\it Chenopodium quinoa}, \cite{jarvis2017genome})  
and other HMM-based tools for many more (e.g. {\it Citrus clementina}, \cite{wu2018genomics}; and
{\it Capsella grandiflora}, \cite{slotte2013capsella}). The Ensemble animal dataset by and large
used the Ensembl annotation pipeline, which includes no HMM-based tool.

We downloaded and processed public RNAseq data to obtain an independent option 
for evaluating model performance. We selected three each of plant and animal genomes for
detailed evaluation with RNAseq. These were selected to have relatively good 
({\it Manihot esculenta} and {\it Papio anubis}), 
typical ({\it Medicago truncatula} and {\it Equus caballus}), 
and poor ({\it Theobroma cacao} and {\it Petromyzon marinus})
performance compared to Augustus \ref{supfig:rna_select} % subgenic F1 sorted by Helixer new plots
within our generalizable range (i.e. excluding the outgroups algae and
invertebrates). Selections were further constrained by the availability of stranded RNAseq
data.

% todo, add versions to everything below
For each of these six species the following search was performed on Sequence Read Archive
'(("<species name>"[Organism] OR <species name>[All Fields]) AND stranded[All Fields]) 
AND ("biomol rna"[Properties] AND "library layout paired"[Properties])'. In so far as 
more than 50 samples were identified, every Nth sample was selected so that about 50
samples were chosen for further processing, as listed here \ref{}. % todo make a supp table
Each sample was quality controlled with FastQC (www.bioinformatics.babraham.ac.uk/projects/fastqc/), before and after trimming
with Trimmomatic \citep{bolger2014trimmomatic}, and then mapped to the genome with Hisat2 \citep{kim2019graph}. 
Samtools \citep{li2009sequence} was used for the sorting and indexing that was necessary both for
RNAseq mapped metric collection with PicardTools \citep{Picard2019toolkit} and custom downstream 
processing below. MultiQC \citep{ewels2016multiqc} was used to summarize the results. The above was 
automated and the results visualized with the code here (... make available and link to RNAseq stuff).
Samples were filtered to those that had {\it relatively} high mapping rates, high mapping to
exonic regions relative to other regions; that had low 3' bias, normal FastQC and 
Trimmomatic stats, and that
were confirmed to be stranded with 2nd strand as sense strand. If more remained, seven
of the high quality samples were selected randomly.

@@

\begin{figure*}[!tpb]
\label{fig:main_results}
\centerline{\includegraphics[width=\textwidth]{images/main_results}}
\caption{Main overall results and other investigations. The top row shows the animal
 data; the bottom row is concerned with the plants. Orange dots are for AUGUSTUS and 
the blue ones are for Helixer. {\bf a)} Swarm - and boxplots showing the Subgenic F1 
scores for all genomes that were not used during training of Helixer, which is 
equivalent to all validation and test genomes. Each dot represents the prediction 
performace for one genome. The median prediction scores are shown by the red line. 
Details of how AUGUSTUS was used are given in Section \ref{sec:augustus}. {\bf b)} 
Scatterplots showing the prediction performances measured in Subgenic F1 by genome 
length. The regression line is shown with a 95\% confidence interval. {\bf c)} The 
average Genic F1 (in red) and basepair wise accuracy (in purple) with respect to 
the position in the 20000 basepair long input sequences. The dashed line shows the 
same without overlapping. Each value is the average performance of a 200 basepair 
long subsection by one of the models of the final ensemble. 
See Section \ref{supsec:overlapping} of the 
Supplements for the effects of overlapping in individual species.}
\end{figure*}

@@

\begin{figure*}[!h]
\label{fig:cov_example_main}
\centerline{\includegraphics[width=\textwidth]{images/cov_examples/cov_example_main}}
\caption{Four example helixer predictions in the context of RNAseq data, the reference,
and Augustus' prediction for {\it M. esculenta}. The examples were chosen so
that a) Helixer had high accuracy against the reference and the reference
was supported by the RNAseq data, b) Helixer had high accuracy but the 
reference was not supported by the RNAseq data, c) Helixer had low accuracy
and the reference was supported by the RNAseq data, and d) helixer had low
accuracy but the reference was not supported by RNAseq data. Plotability was
a major secondary consideration. Each subplot shows from top to
bottom i) the natural log of the coverage (``cov", solid) and spliced coverage 
(``sc", dotted) + 1, ii) the reference annotation in matrix form, iii) 
Augustus' predictions in matrix form, and iv) Helixer's predictions. The reference
and Augustus have either 0 (white) or 1 (black) for each base pair and category, while
Helixer emits a probability from 0-1 represented via gray-scale. ``Ntrn" stand
for intron, and ``IG" stands for intergenic.
}
\end{figure*}

As an independent evaluation and to get a better qualitative understanding of
performance we compared Helixer predictions, Augustus predictions and the
reference to RNAseq coverage data. RNAseq was processed for three each of plants
and animals chosen because they had varied performance quality relative to Augustus. 
RNAseq coverage provides support for an 
exonic annotation (UTR or CDS), spliced coverages provides support for an 
annotation of intron, and neither coverage nor spliced coverage is expected 
for intergenic annotations. Looking at selected subsequences (Figure \ref{fig:cov_example_main}) % and supplemental figures.
we identified cases where RNAseq supported both the annotations of the reference 
and Helixer a), neither b), the reference, but not Helixer c),
and Helixer, but not the reference d).

@@

Helixer models also sometimes showed uncertainty for larger regions.
In some cases, where Helixer did not receive RNAseq support for its highest probability
annotation, it assigned a low but non-trivial probability to the RNAseq-supported 
exon/intron pattern (Figures \ref{fig:cov_example_main}b-c,S\ref{supfig:cov_example_02}c). However, in the extreme,
there are cases where the Helixer model exibits substantial indecision or confusion 
and shifts gradually between classes with no single class receiving a high probability 
for extended stretches (Figures S\ref{supfig:cov_example_02}c, S\ref{supfig:cov_example_02}c-d). 
Notably, in one of the examples (S\ref{supfig:cov_example_02}d))
the RNAseq shows evidence of alternative splicing; 
and in another (S\ref{supfig:cov_example_02}c) Helixer's
prediction falls between that of the reference and Augustus.

\begin{figure*}[!h]
\label{fig:average_vs_augustus}
\includegraphics[width=8.6cm]{images/cov_examples/average_vs_augustus}  % should be 8.6 when floats are sorted out right
\caption{Fraction of bp with color-indicated a) coverage and b) spliced coverage of genomic positions 
broken down by the confusion matrix of Augustus' and Helixer's predictions. Categories 
are only displayed if they can be meaningfully compared by examining a) coverage or b) spliced
coverage. The displayed fractions are the averages of the individual fractions for the 
six RNAseq-evaluation species. The left-most bars show cases where the two tools agree, 
while the remaining bars show paired conflicts. ``Ntrn" stand
for intron, and ``IG" stands for intergenic.}
\end{figure*}

Coverage and Spliced coverage were broken down by the confusion matrix (of both
Helixer vs Augustus and Helixer vs the reference) for all genomes (Figure \ref{fig:average_vs_augustus}). % H vs A main, H vs R supp
Where tools were in agreement (far left), coverage and spliced coverage closely matched
expectations. Specifically for CDS:CDS and UTR:UTR most base pairs had some, and many
had moderate or high coverage. The same pattern was seen for intron:intron and spliced
coverage. Finally, intergenic:intergenic showed only a small fraction of base pairs with 
any of either coverage or spliced coverage. In all conflicts the amount of RNAseq support
fell between the cases where tools agreed, indicating that all options were at least
capable of finding weak-spots in the other annotations.
In cases of conflict, Helixer's predictions were substantially more consistent with
the RNAseq data than those of Augustus. Specifically, for base pairs where one tool said CDS and
the other intergenic or intron, there was more coverage when 
Helixer said CDS (Figure \ref{fig:average_vs_augustus}a). Similarly, for base pairs where one tool said intron and the other 
CDS or intergenic, there was more spliced coverage when Helixer said intron (Figure \ref{fig:average_vs_augustus}b). 
These patterns were consistent in direction but varied in magnitude in the individual species,
with the expception of {\it P. marinus} where Helixer and Augustus performed comparably
at differentiating CDS and introns (Figure S\ref{supfig:each_sp_vs_augustus}).

RNAseq based comparison of Helixer and the reference was less clear cut.
Averaged across species, helixer's predictions received slightly more support
than the reference when differentiating CDS, UTR and introns from intergenic;
however helixer's predictions received slightly less support than the reference 
when differentiating CDS and UTR from introns (Figure S\ref{supfig:average_vs_ref}). Performance varies between individual
species, from {\it P. marinus}, where the reference receives more support
in every conflict, to {\it T. cacao}, where Helixer models receive equivalent
or more support in every conflict (Figure S\ref{supfig:each_sp_vs_ref}). Interestingly, these are the two species that
were selected because they had (relative to the usual difference) low Subgenic F1 compared to 
Augustus; for the former the RNAseq confirms relatively weak performance for Helixer, while the 
for the latter RNAseq rather indicates a sub-par reference.

@@

Finally, developent of a post-processing method
to go from base pair wise predictions to integrated predictions for whole 
transcripts at each loci could both further improve performance and would
greatly increase real world applications.
\FloatBarrier

@@

\bibliographystyle{natbib}
\bibliography{literature}
% place holders for supplemental things
% ori version was grepped out of supplements, then
% :'<,'>s/^/\\begin{supfig}/g
% :'<,'>s/$/\\end{supfig}/g
% and similar for tables/sections
% tables
\begin{suptab}\label{suptab:generalization_animals}\end{suptab}
\begin{suptab}\label{suptab:generalization_plants}\end{suptab}
\begin{suptab}\label{suptab:params}\end{suptab}
\begin{suptab}\label{suptab:classweights}\end{suptab}
\begin{suptab}\label{suptab:prediction_lengths}\end{suptab}
\begin{suptab}\label{suptab:genome_stats}\end{suptab}
\begin{suptab}\label{suptab:dcnn_xval}\end{suptab}
\begin{suptab}\label{suptab:dcnn_params}\end{suptab}
% figures
\begin{supfig}\label{supfig:phylogeny_f1_animals}\end{supfig}
\begin{supfig}\label{supfig:phylogeny_differences_animals}\end{supfig}
\begin{supfig}\label{supfig:phylogeny_f1_plants}\end{supfig}
\begin{supfig}\label{supfig:phylogeny_differences_plants}\end{supfig}
\begin{supfig}\label{supfig:overlapping_animals}\end{supfig}
\begin{supfig}\label{supfig:overlapping_plants}\end{supfig}
\begin{supfig}\label{supfig:cov_example_02}\end{supfig}
\begin{supfig}\label{supfig:cov_example_03}\end{supfig}
\begin{supfig}\label{supfig:UTRs_animals}\end{supfig}
\begin{supfig}\label{supfig:UTRs_plants}\end{supfig}
\begin{supfig}\label{supfig:average_vs_ref}\end{supfig}
\begin{supfig}\label{supfig:each_sp_vs_augustus}\end{supfig}
\begin{supfig}\label{supfig:each_sp_vs_ref}\end{supfig}
% sections
\begin{supsec}\label{sec:generalization}\end{supsec}
\begin{supsec}\label{sec:hyperparameters}\end{supsec}
\begin{supsec}\label{sec:longer}\end{supsec}
\begin{supsec}\label{sec:overlapping}\end{supsec}
\begin{supsec}\label{sec:training_data}\end{supsec}
\begin{supsec}\label{sec:dcnn}\end{supsec}
\begin{supsec}\label{sec:rnaseq_eval}\end{supsec}

>>>>

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Manuscript Category}
\usepackage{url}

@@

\subsection{Data Pre-processing}
We first pre-process and store the raw genomic information by using GeenuFF (https://https:
//github.com/weberlab-hhu/GeenuFF). GeenuFF is a tool for checking and exploring genomic data, that 
stores all information inside a SQL database. Training and evaluation-ready data is 
generated by querying this database and then transforming the returned data into a 
numerical format suitable for machine learning. The encoding of the genomic sequence 
is done in line with the IUPAC nucleic acid notation and the functional annotation 
used as labels during training is transformed to a one hot encoding with the four 
classes {\it Intergenic}, {\it UTR}, {\it CDS} and {\it Intron}. See also Table 
\ref{tab:arrays} for a more detailed description of the generated data types.

@@

Our final model was trained with 10 bases of genomic sequence as input during each 
time step and produces individual predictions for each of those 10 bases 
simultaneously. This grouping enabled us to train effectively with far longer 
sequences than usual. 
We compare our final models to a dilated CNN (dCNN) baseline as this kind of architecture is 
widely used when working with DNA data as input. Details of the neural architecture 
search are given in Section \ref{ssec:dcnn} of the Supplements.

\subsection{Inference Techniques}
We also used multiple techniques to improve the prediction quality after the 
training was done. One very effective way for genomes with larger genes was to input 
longer sequences during inference time than during training. This was done for all 
animal genomes except the invertebrates. The input sequences were up to 10 times 
longer, depending on the phylogenetic group and assembly quality. This also demonstrates the 
ability of our model to generalize as it is able to make successful prediction on 
far longer sequences than it has ever seen. The 
concrete lengths were chosen to keep the typical average gene length roughly 
proportional to the length of the sequence input. For more implementation details 
on this see Section \ref{ssec:longer} of the Supplements or the source code.

The final predictions of a single model are constructed by overlapping predictions, 
that were made from a sliding window and then cropped to a core sequence. This was done 
to strongly reduce a typical drop in performance of the models towards the beginning 
and end of each sequence (see Figure \ref{fig:main_results}c). It also improves the 
average model performance in general by providing the model with multiple different 
starting points. The different overlapping sequences were combined by averaging 
the individual softmax values of each base. The figures in Section \ref{ssec:overlapping} of the 
Supplements show the effect of overlapping for each genome, ordered by N75. We found 
that overlapping tends to work best if the genomes are not very fragmented and we 
used it for both animals and plants.

@@

\subsection{Evaluation of AUGUSTUS} 
\label{sec:augustus}
Due to the practical time constraints for retraining and running, we decided to compare our Helixer
models only to one existing {\it de novo} gene caller, the popular gene caller, AUGUSTUS \citep{stanke2003gene}.
Typical usage requires retraining AUGUSTUS for each species, with the exception of a
few lucky cases where a model for a sufficiently close relative is already available. To scale this for 
the large plant and animal datasets, we used protein-homology to create a training set and
therefore could train and evaluate only models without UTRs.

Orthologs of highly conserved generally single-copy genes were identified in each genome using
BUSCO. The {\it viridiplantae} set was used for plants and the {\it metazoa} set for animals.
The training genbank files were generated directly by BUSCO, by utilizing the ``--long"  % inline code?
parameter. For plants, the entire retraining could be performed as above; however, for animals
we randomly selected only half the BUSCO-generated training set, which resulted in a training set 
size and runtime comparable to the plants (~400 genes and several days per species). For animals
the training was carried out with the subsetted training file and using the following scripts provided
by AUGUSTUS. An untrained model was setup with "new\_species.pl", and the model was fit by running
"etraining" before and after the major hyper-parameter optimization with "optimize\_augustus.pl".
Using the trained model for each species, we ran the main prediction ("augustus") with ``--UTR=off"
and ``--gff3=on". The gffs produced by AUGUSTUS were imported into GeenuFF and exported as h5
files in the same manor as the reference, allowing direct comparison with reference and 
Helixer predictions. While tenable for some 237 species, this method allowed neither training
nor prediction of UTR regions with AUGUSTUS, so the metric Subgenic F1, which excludes UTR,
was used for all comparisons between AUGUSTUS and Helixer.

\subsection{Evaluation against independent RNAseq data}
\label{sec:rnaseq}
The reference annotations were created with existing tools and largely with a pipeline incorporating
{\it de novo} gene predictions with RNAseq and homology data. The references, like any
data, are expected to contain errors so we chose to use RNAseq data for independent evaluation.
In plants there is the additional concern that the references may share biases with
AUGUSTUS or HMMs as AUGUSTUS was used as the {\it de novo} gene caller for the reference of
many plant species % (e.g. {\it Arabidopsis lyrata}, \cite{rawat2015improving}; 
%{\it Asparagus officinalis}, 
%\cite{harkess2017asparagus}; and {\it Chenopodium quinoa}, \cite{jarvis2017genome})  
and other HMM-based tools for many more %(e.g. {\it Citrus clementina}, \cite{wu2018genomics}; and
%{\it Capsella grandiflora}, \cite{slotte2013capsella}). 
The Ensemble animal dataset by and large
used the Ensembl annotation pipeline, which includes no HMM-based tool.

We downloaded and processed public RNAseq data to obtain an independent option 
for evaluating model performance. We selected three each of plant and animal genomes for
detailed evaluation with RNAseq. These were selected to have relatively good 
({\it Manihot esculenta} and {\it Papio anubis}), 
typical ({\it Medicago truncatula} and {\it Equus caballus}), 
and poor ({\it Theobroma cacao} and {\it Petromyzon marinus})
performance compared to AUGUSTUS (Figure S\ref{supfig:rna_select_animals}-\ref{supfig:rna_select_plants}) % subgenic F1 sorted by Helixer new plots
within our generalizable range (i.e. excluding the outgroups algae and
invertebrates). Selections were further constrained by the availability of stranded RNAseq
data.

% todo, add versions to everything below
For each of these six species the following search was performed on Sequence Read Archive
`(("<species name>"[Organism] OR <species name>[All Fields]) AND stranded[All Fields]) 
AND ("biomol rna"[Properties] AND "library layout paired"[Properties])'. In so far as 
more than 50 samples were identified, every Nth sample was selected so that under 50
samples were chosen for further processing (Table S\ref{suptab:rnaseq_samples}).
Each sample was quality controlled with FastQC (www.bioinformatics.babraham.ac.uk/projects/fastqc/), before and after trimming
with Trimmomatic \citep{bolger2014trimmomatic}, and then mapped to the genome with Hisat2 \citep{kim2019graph}. 
Samtools \citep{li2009sequence} was used for the sorting and indexing that was necessary both for
RNAseq mapped metric collection with PicardTools (http://broadinstitute.github.io/picard/) and custom downstream 
processing below. MultiQC \citep{ewels2016multiqc} was used to summarize the results. The above was 
automated and the results visualized with the code here (... make available and link to RNAseq stuff).
Samples were filtered to those that had {\it relatively} high mapping rates, high mapping to
exonic regions relative to other regions; that had low 3' bias, normal FastQC and 
Trimmomatic stats, and that
were confirmed to be stranded with 2nd strand as sense strand. If more remained, seven
of the high quality samples were selected randomly.

@@

\begin{figure*}[!tpb]
\label{fig:main_results}
\centerline{\includegraphics[width=\textwidth]{images/main_results}}
\caption{Main overall results and other investigations. The top row shows the animal
 data; the bottom row is concerned with the plants. Orange dots are for AUGUSTUS and 
the blue ones are for Helixer. {\bf a)} Swarm - and boxplots showing the Subgenic F1 
scores for all genomes that were not used during training of Helixer, which is 
equivalent to all validation and test genomes. Each dot represents the prediction 
performace for one genome. The median prediction scores are shown by the red line. 
Details of how AUGUSTUS was used are given in Section \ref{sec:augustus}. {\bf b)} 
Scatterplots showing the prediction performances measured in Subgenic F1 by genome 
length. The regression line is shown with a 95\% confidence interval. {\bf c)} The 
average Genic F1 (in red) and basepair wise accuracy (in purple) with respect to 
the position in the 20000 basepair long input sequences. The dashed line shows the 
same without overlapping. Each value is the average performance of a 200 basepair 
long subsection by one of the models of the final ensemble. 
See Section \ref{ssec:overlapping} of the 
Supplements for the effects of overlapping in individual species.}
\end{figure*}

@@

\begin{figure*}[!h]
\label{fig:cov_example_main}
\centerline{\includegraphics[width=\textwidth]{images/cov_examples/cov_example_main}}
\caption{Four example helixer predictions in the context of RNAseq data, the reference,
and AUGUSTUS' prediction for {\it M. esculenta}. The examples were chosen so
that a) Helixer had high accuracy against the reference and the reference
was supported by the RNAseq data, b) Helixer had high accuracy but the 
reference was not supported by the RNAseq data, c) Helixer had low accuracy
and the reference was supported by the RNAseq data, and d) helixer had low
accuracy but the reference was not supported by RNAseq data. Plotability was
a major secondary consideration. Each subplot shows from top to
bottom i) the natural log of the coverage (``cov", solid) and spliced coverage 
(``sc", dotted) + 1, ii) the reference annotation in matrix form, iii) 
AUGUSTUS' predictions in matrix form, and iv) Helixer's predictions. The reference
and AUGUSTUS have either 0 (white) or 1 (black) for each base pair and category, while
Helixer emits a probability from 0-1 represented via gray-scale. ``Ntrn" stand
for intron, and ``IG" stands for intergenic.
}
\end{figure*}

As an independent evaluation and to get a better qualitative understanding of
performance we compared Helixer predictions, AUGUSTUS predictions and the
reference to RNAseq coverage data. RNAseq was processed for three each of plants
and animals chosen because they had varied performance quality relative to AUGUSTUS. 
RNAseq coverage provides support for an 
exonic annotation (UTR or CDS), spliced coverages provides support for an 
annotation of intron, and neither coverage nor spliced coverage is expected 
for intergenic annotations. Looking at selected subsequences (Figure \ref{fig:cov_example_main}) % and supplemental figures.
we identified cases where RNAseq supported both the annotations of the reference 
and Helixer a), neither b), the reference, but not Helixer c),
and Helixer, but not the reference d).

@@

Helixer models also sometimes showed uncertainty for larger regions.
In some cases, where Helixer did not receive RNAseq support for its highest probability
annotation, it assigned a low but non-trivial probability to the RNAseq-supported 
exon/intron pattern (Figures \ref{fig:cov_example_main}b-c,S\ref{supfig:cov_example_02}c). However, in the extreme,
there are cases where the Helixer model exibits substantial indecision or confusion 
and shifts gradually between classes with no single class receiving a high probability 
for extended stretches (Figures S\ref{supfig:cov_example_02}c, S\ref{supfig:cov_example_02}c-d). 
Notably, in one of the examples (S\ref{supfig:cov_example_02}d))
the RNAseq shows evidence of alternative splicing; 
and in another (S\ref{supfig:cov_example_02}c) Helixer's
prediction falls between that of the reference and AUGUSTUS.

\begin{figure}[hbt]
\label{fig:average_vs_augustus}
\includegraphics[width=8.6cm]{images/cov_examples/average_vs_augustus}  % should be 8.6 when floats are sorted out right
\caption{Fraction of bp with color-indicated a) coverage and b) spliced coverage of genomic positions 
broken down by the confusion matrix of AUGUSTUS' and Helixer's predictions. Categories 
are only displayed if they can be meaningfully compared by examining a) coverage or b) spliced
coverage. The displayed fractions are the averages of the individual fractions for the 
six RNAseq-evaluation species. The left-most bars show cases where the two tools agree, 
while the remaining bars show paired conflicts. ``Ntrn" stand
for intron, and ``IG" stands for intergenic.}
\end{figure}

Coverage and Spliced coverage were broken down by the confusion matrix (of both
Helixer vs AUGUSTUS and Helixer vs the reference) for all genomes (Figure \ref{fig:average_vs_augustus}). % H vs A main, H vs R supp
Where tools were in agreement (far left), coverage and spliced coverage closely matched
expectations. Specifically for CDS:CDS and UTR:UTR most base pairs had some, and many
had moderate or high coverage. The same pattern was seen for intron:intron and spliced
coverage. Finally, intergenic:intergenic showed only a small fraction of base pairs with 
any of either coverage or spliced coverage. In all conflicts the amount of RNAseq support
fell between the cases where tools agreed, indicating that all options were at least
capable of finding weak-spots in the other annotations.
In cases of conflict, Helixer's predictions were substantially more consistent with
the RNAseq data than those of AUGUSTUS. Specifically, for base pairs where one tool said CDS and
the other intergenic or intron, there was more coverage when 
Helixer said CDS (Figure \ref{fig:average_vs_augustus}a). Similarly, for base pairs where one tool said intron and the other 
CDS or intergenic, there was more spliced coverage when Helixer said intron (Figure \ref{fig:average_vs_augustus}b). 
These patterns were consistent in direction but varied in magnitude in the individual species,
with the expception of {\it P. marinus} where Helixer and AUGUSTUS performed comparably
at differentiating CDS and introns (Figure S\ref{supfig:each_sp_vs_augustus}).

RNAseq based comparison of Helixer and the reference was less clear cut.
Averaged across species, helixer's predictions received slightly more support
than the reference when differentiating CDS, UTR and introns from intergenic;
however helixer's predictions received slightly less support than the reference 
when differentiating CDS and UTR from introns (Figure S\ref{supfig:average_vs_ref}). Performance varies between individual
species, from {\it P. marinus}, where the reference receives more support
in every conflict, to {\it T. cacao}, where Helixer models receive equivalent
or more support in every conflict (Figure S\ref{supfig:each_sp_vs_ref}). Interestingly, these are the two species that
were selected because they had (relative to the usual difference) low Subgenic F1 compared to 
AUGUSTUS; for the former the RNAseq confirms relatively weak performance for Helixer, while the 
for the latter RNAseq rather indicates a sub-par reference.

@@

Finally, developent of a post-processing method
to go from base pair wise predictions to integrated predictions for whole 
transcripts at each loci could both further improve performance and would
greatly increase real world applications.

@@

\bibliographystyle{natbib}
\bibliography{literature}
% place holders for supplemental things
% ori version was grepped out of supplements, then
% :'<,'>s/^/\\begin{supfig}/g
% :'<,'>s/$/\\end{supfig}/g
% and similar for tables/sections
% tables
\begin{suptab}\label{suptab:rnaseq_samples}\end{suptab}
\begin{suptab}\label{suptab:generalization_animals}\end{suptab}
\begin{suptab}\label{suptab:generalization_plants}\end{suptab}
\begin{suptab}\label{suptab:params}\end{suptab}
\begin{suptab}\label{suptab:classweights}\end{suptab}
\begin{suptab}\label{suptab:prediction_lengths}\end{suptab}
\begin{suptab}\label{suptab:genome_stats}\end{suptab}
\begin{suptab}\label{suptab:dcnn_xval}\end{suptab}
\begin{suptab}\label{suptab:dcnn_params}\end{suptab}
% figures
\begin{supfig}\label{supfig:rna_select_animals}\end{supfig}
\begin{supfig}\label{supfig:rna_select_plants}\end{supfig}
\begin{supfig}\label{supfig:phylogeny_f1_animals}\end{supfig}
\begin{supfig}\label{supfig:phylogeny_differences_animals}\end{supfig}
\begin{supfig}\label{supfig:phylogeny_f1_plants}\end{supfig}
\begin{supfig}\label{supfig:phylogeny_differences_plants}\end{supfig}
\begin{supfig}\label{supfig:overlapping_animals}\end{supfig}
\begin{supfig}\label{supfig:overlapping_plants}\end{supfig}
\begin{supfig}\label{supfig:cov_example_02}\end{supfig}
\begin{supfig}\label{supfig:cov_example_03}\end{supfig}
\begin{supfig}\label{supfig:UTRs_animals}\end{supfig}
\begin{supfig}\label{supfig:UTRs_plants}\end{supfig}
\begin{supfig}\label{supfig:average_vs_ref}\end{supfig}
\begin{supfig}\label{supfig:each_sp_vs_augustus}\end{supfig}
\begin{supfig}\label{supfig:each_sp_vs_ref}\end{supfig}
% sections
\begin{supsec}\label{ssec:data_prep}\end{supsec}
\begin{supsec}\label{ssec:generalization}\end{supsec}
\begin{supsec}\label{ssec:hyperparameters}\end{supsec}
\begin{supsec}\label{ssec:longer}\end{supsec}
\begin{supsec}\label{ssec:overlapping}\end{supsec}
\begin{supsec}\label{ssec:training_data}\end{supsec}
\begin{supsec}\label{ssec:dcnn}\end{supsec}
\begin{supsec}\label{ssec:rnaseq_eval}\end{supsec}
